---
title: "date4ts"
format: html
editor: visual
---

## **How to Use date4ts: A Date Wrangling Package for Time Series Analysis in R**

## The date4ts package offers precise, deliberate control over date formatting, merging, and extraction for time series workflows

Time series analysis often runs smoothly until dates become a problem. A retail analyst building monthly revenue reports might import data from two separate systems — one that outputs standard R *Date* objects and another that uses decimal year formats like 2023.25. Reconciling those two date representations into a single cohesive *ts* object can cost more time than the analysis itself.

The **date4ts** package addresses exactly this friction. Released on CRAN and compatible with R 4.1 and above, date4ts provides a focused set of utility functions for converting, merging, modifying, and navigating dates within standard *ts* objects. It covers the tasks that sit between "load the data" and "run the model" — the wrangling layer that every time series workflow needs, but few packages explicitly support.

## **Installing date4ts**

Getting started requires the standard installation call from CRAN (or from the Packages tab in RStudio — you can search and download the package, then call it in your script via the *library()* function.

```{r}
install.packages("date4ts")
library(date4ts)
```

## **Formatting and Converting Dates**

The most immediate problem date4ts solves is translation between R’s native date formats and the *c(YYYY, period)* vector format that the *ts()* function expects.

## **From Date Objects to Time Series Format**

The *date2date_ts()* function accepts a standard R *Date* object and a frequency argument — either 12 for monthly or 4 for quarterly — and returns the *c(YYYY, MM/TT)* vector that *ts()* uses to anchor a series.

```{r}
## Convert a standard Date to time series format
my_date <- as.Date("205-05-05")

## date2date_ts converts a Date into a c(year, period) vector
ts_date <- date2date_ts(my_date, frequency = 12)
```

This conversion matters because *ts()* itself does not parse *Date* objects directly; it requires this structured vector as its start argument. Having a dedicated function for that translation eliminates a common source of indexing errors, particularly when the starting period is not January or Q1.

## **From Decimal Years to Structured Vectors**

Decimal year formats like *2022.25* appear frequently in exported data from econometric software, financial APIs, and legacy systems. The *as_yyyytt()* and *as_yyyymm()* functions convert those values into structured date vectors.

```{r}
## Convert a decimal year TimeUnit into a structured date vector
decimal_year <- 2022.25
```

Here is how the as_yyyytt() function operates…

```{r}
## as_yyyytt returns the period in year-quarter format
qtr_date <- as_yyyytt(decimal_year)
qtr_date
```

…and now here is the as_yyyymm() function and its yield.

```{r}
month_date <- as_yyyymm(decimal_year)
month_date
```

The *as_yyyytt()* functionis the more general converter, returning whatever period-unit format corresponds to the given frequency. The *as_yyyymm()* function is the monthly-specific variant, making the intent of the code explicit when working in month-level data. Both functions share the same underlying logic, so choosing between them is primarily a matter of communicating your frequency to colleagues reading your code.

## **Reversing the Conversion**

The reverse path — from a time series position back to a standard *Date* — is handled by the function *date_ts2date()*.

```{r}
## Both arguments must be explicit integers - use the L suffix
ts_position <- c(2023L, 4L)

## date_ts2date reconstructs the Date from year and period
back_to_date <- date_ts2date(date_ts = ts_position, frequency_ts = 12L)
back_to_date
```

The *date_ts2date()* function completes the round-trip conversion, returning a standard `Date` object from a time series position vector. Both the *date_ts* and *frequency_ts* arguments expect explicit R integers — written with the L suffix (c(2023L, 4L) and 12L — because the package uses **checkmate** for internal validation and will warn when plain doubles are passed instead. This is particularly useful when passing time series results into packages like **ggplot2** or **dplyr** that expect standard date columns. Rather than manually reconstructing months from integer period values, a single function call handles the arithmetic correctly across year boundaries.

```{r}
## date_ts2date reconstructs the Date from year and period
back_to_date <- date_ts2date(date_ts = ts_position, frequency_ts = 12L)
back_to_date
## [1] "2023-04-01"

## Create two overlapping monthly ts objects
sales_2022 <- ts(c(120, 135, 110, 145, 160, 175), start = c(2022, 1), frequency = 12)
sales_new  <- ts(c(178, 190, 205, 215), start = c(2022, 5), frequency = 12)

## combine2ts merges both; where they overlap, b takes priority
combined_sales <- combine2ts(sales_2022, sales_new)
combined_sales

## Extend an existing ts with new end-of-period values
base_series <- ts(c(500, 520, 515, 530, 545), start = c(2023, 1), frequency = 12)

## extend_ts appends new values to the end of the series
extended_series <- extend_ts(series = base_series, replacement = c(560, 575, 590))
extended_series
```

```{r}
## Correct a single value at a known date position
corrected_series <- set_value_ts(
  series      = extended_series,
  date_ts     = c(2023L, 3L),  ## integer L suffix required
  replacement = 518
)

corrected_series
```

```{r}
# Getting values by Data Range
## Extract a range of values from a monthly ts
quarterly_window <- get_value_ts(
  series    = combined_sales,
  date_from = c(2022L, 3L),
  date_to   = c(2022L, 6L)
)
quarterly_window

## Find the next and previous period relative to a given date
current_period <- c(2023L, 6L)  ## June 2023
```

```{r}
## next_date_ts advances one period forward
next_period <- next_date_ts(date_ts = current_period, frequency_ts = 12L)
next_period
## [1] 2023    7

## previous_date_ts steps one period back
prev_period <- previous_date_ts(date_ts = current_period, frequency_ts = 12L)
prev_period

## Identify the boundaries of actual data within a padded series
padded_series <- ts(
  c(NA, NA, 300, 320, 335, 350, NA),
  start = c(2023, 1),
  frequency = 12
)

## first_date returns the first period with a non-NA value
data_start <- first_date(padded_series)
data_start
## [1] 2023    3

## last_date returns the last period with a non-NA value
data_end <- last_date(padded_series)
data_end
## [1] 2023    6
```

```{r}
## Build a sample quarterly ts for visualization
quarterly_revenue <- ts(
  c(1200, 1350, 1100, 1450, 1600, 1750, 1820, 1900),
  start = c(2022, 1),
  frequency = 4
)

## ts2df converts the ts object into a tidy data frame
revenue_df <- ts2df(quarterly_revenue)
revenue_df

##    year quarter value
## 1  2022       1  1200
## 2  2022       2  1350
## 3  2022       3  1100
## 4  2022       4  1450
## 5  2023       1  1600
## 6  2023       2  1750
## 7  2023       3  1820
## 8  2023       4  1900
```

```{r}
## Now the data is ready for ggplot2, dplyr, or gt
library(ggplot2)

revenue_df |>
  ggplot(aes(x = date, y = x, group = 1)) +
  geom_line(color = "#2A7F8E", linewidth = 1.2) +
  geom_point(color = "#1B3A5C", size = 3) +
  labs(
    title    = "Quarterly Revenue Trend",
    subtitle = "Converted from ts object using ts2df()",
    x        = "Period",
    y        = "Revenue"
  ) +
  theme_minimal()
```
