---
title: "Donut Chart"
format: html
editor: visual
---

# **Donut Chart with R Programming Language**

### **Step-by-Step Donut Chart**

```{r}
library(tidyverse)
veriler <- c(100, 300, 250, 400, 600)
data <- data.frame(
    xxx = LETTERS[1:length(veriler)],  
    deger = veriler
)
data <- data %>%
    mutate(
        fraction = deger / sum(deger),
        ymax = cumsum(fraction),
        ymin = c(0, head(ymax, -1))
    )
ggplot(data, aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 3, fill = xxx)) +  
    geom_rect(color = "black") +
    coord_polar(theta = "y") +
    xlim(c(2, 4)) +
    theme_void() +
    labs(title = "ABCDE") +
    scale_fill_brewer(palette = "Set2")
```

Let’s explain in stages:

-   `geom_rect()` → Draws each slice as a rectangle.

-   `coord_polar(theta = "y")` → Transforms rectangles into circular form, creating a "pie chart" appearance.

-   `xlim(c(2, 4))` → Leaves space in the middle of the chart; this space forms the donut's center.

-   `theme_void()` → Removes unnecessary elements like axes, grids, and text.

-   `labs(title = "ABCDE")` → Adds a title to the chart.

-   `scale_fill_brewer(palette = "Set3")` → Determines the color palette, ensuring visual harmony.

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
