---
title: "Drawing ınteractive maps"
format: html
editor: visual
---

# **Drawing ınteractive maps with R: from scratch to user presentation**

By the end of the notebook, you will have produced a map with the following features:

-   Markers showing city locations

-   Marker clustering

-   HTML popup windows

-   Satellite / standard map switching

-   Mini-map

-   Scale bar

-   Drawing tools on the map

And all of this with a single R script.

-   Working Directory

R wants to know which folder it’s working in.

getwd()

### Install packages

```{r}
install.packages("leaflet")
install.packages("leaflet.extras")
```

### Library

```{r}
library(leaflet)
library(leaflet.extras)
```

### Interactive maps

```{r}
library(leaflet)
library(leaflet.extras)

# 1. DATA: Added your new locations
cities <- data.frame(
  name = c("London", "Manchester", "Birmingham", "Aberdeen", "Bristol", 
           "Glasgow", "Edinburgh", "Bath", "Cardiff", "Derry", "Oban", 
           "Orkney Mainland", "North Ronaldsay", "Papa Westray", "Mallaig", 
           "Brighton", "Belfast", "Portland"),
  lat  = c(51.5074, 53.4808, 52.4862, 57.1497, 51.4545, 
           55.8642, 55.9533, 51.3758, 51.4816, 54.9966, 56.4127, 
           58.9809, 59.3667, 59.3500, 57.0065, 
           50.8225, 54.5973, 50.5471),
  lng  = c(-0.1278, -2.2426, -1.8904, -2.0943, -2.5879, 
           -4.2518, -3.1883, -2.3599, -3.1791, -7.3086, -5.4706, 
           -3.1360, -2.4333, -2.9000, -5.8306, 
           -0.1372, -5.9301, -2.4411)
)

# 2. HTML POPUP CONTENT
cities$popup_html <- paste0(
  "<b style='font-size:14px;'>", cities$name, "</b><br>",
  "Latitude: ", cities$lat, "<br>",
  "Longitude: ", cities$lng
)

# 3. CUSTOM ICON FIX
# Note: Removed system.file(). Use the direct path to your image.
# Define your image dimensions first
# Define your specific image dimensions (Check your file properties for exact pixels)
# Let's assume your pinSmiley01.png is 40 pixels wide and 60 pixels high
my_w <- 40 
my_h <- 40

default_icon <- makeIcon(
  iconUrl = "/Users/ulrike_imac_air/projects/maps/MyTravelMap/pins/Happy_smiley_face.png",
  iconWidth = my_w, 
  iconHeight = my_h,
  
  # THE FIX:
  iconAnchorX = my_w / 2, # Centered horizontally
  iconAnchorY = my_h,     # Tip at the very bottom
  
  # Optional: Keep the popup from covering the pin
  popupAnchorX = 0,
  popupAnchorY = -my_h   # Moves the popup window to the top of the pin
)

# 4. INTERACTIVE MAP
leaflet(cities) %>% 
  addTiles(group = "Standard") %>% 
  addProviderTiles("Esri.WorldImagery", group = "Satellite") %>% 
  addMarkers(
    ~lng, ~lat,
    popup = ~popup_html,
    icon = default_icon, # Using the custom icon here
    clusterOptions = markerClusterOptions()
  ) %>% 
  setView(lng = -3.5, lat = 55, zoom = 6) %>% 
  addScaleBar(position = "bottomleft") %>% 
  addMiniMap(toggleDisplay = TRUE) %>% 
  addDrawToolbar() %>% 
  addLayersControl(
    baseGroups = c("Standard", "Satellite"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### **Real-World Scenario: How ıs this ınteractive map delivered to the user?**

Let’s say someone you’re working with has asked you for an interactive map chart containing specific locations. You’ve created this map using R and RStudio. The code works, the markers appear, zooming in and out functions… everything is fine. But here’s the real question:

How will you send this interactive map to the user?

When you look at the options in the “Export” menu within RStudio, you’ll see two alternatives:

-   Save Image

-   Save as Web Page

However, neither option works the way you think. Let me explain why, step by step:

### **1. You can’t save it with “Save Image” — because interactivity is lost**

This option only takes a screenshot of the map. That means:

-   no zooming in and out,

-   no clickable markers,

-   no pop-ups,

-   zero interactivity.

In short, the interactive feature of your work is completely lost. So if you want to send an interactive map to a user, the “Save Image” option is completely ruled out.

### **2. “Save as Web Page” seems logical at first glance… but it doesn’t work**

This option saves the map as an HTML file. Then you might think:

“Okay, I’ll save the HTML file to my desktop. Then I’ll send this file to the user. When they open it, the map will work.”

But unfortunately, this doesn’t work either. Because the HTML file contains some support files necessary for the map to display:

-   leaflet CSS

-   leaflet JS

-   dependent script files

-   icon links

-   local paths of dependent libraries

These files reference paths on your computer.

Result:

-   the person opening the file on another computer can’t see the map

-   markers don’t load

-   the map box appears empty

-   “dependency not found” warning appears in the console

In other words, the HTML file only works on your computer.

Therefore, “Save as Web Page” is also not suitable for sending to the user.

### **So then, how do you send an interactive map to the user?**

The only correct way to do this is to publish the map on a web server or hosting service.

### **Easiest: Upload via RPubs**

In RStudio, open this path:

File → Publish to RPubs
