---
title: "InteractiveDashboards"
format: html
editor: visual
---

## **Data Visualization with R: Creating Interactive Dashboards**

## My Hands-On Experience Turning Raw Data into Interactive Insights.

When I first started using **R for visualization**, I was amazed at how powerful libraries like **ggplot2**, **plotly**, and **Shiny** were for building interactive dashboards. Unlike static reports, these dashboards allow users to **explore data dynamically**, making insights more actionable.

## **1) Loading Data into R**

I always begin by importing my dataset and inspecting its structure.

```{r}
# Load necessary libraries
library(readr)
library(dplyr)

# Load dataset
df <- read_csv("sales_data.csv")

# Preview data
head(df)
str(df)
summary(df)
```

## **2) Data Cleaning and Transformation**

Before visualizing, I always make sure the data is clean.

```{r}
# Handle missing values
df <- df |> 
  mutate(
    revenue = ifelse(is.na(revenue), mean(revenue, na.rm = TRUE), revenue)
  )

# Create new features
df <- df |> 
  mutate(Profit_Margin = (revenue - cost) / revenue)

# Inspect transformed data
head(df)
```

Creating new features like **profit margin** often adds **business context** to dashboards.

## **3) Basic Visualization with ggplot2**

I rely heavily on `ggplot2` for static but **highly customizable plots**.

```{r}
library(ggplot2)

# Revenue distribution
ggplot(df, aes(x = revenue)) +
  geom_histogram(binwidth = 5000, fill = "steelblue", color = "black") +
  labs(title = "Revenue Distribution", x = "Revenue", y = "Count")

# Revenue by region
ggplot(df, aes(x = region, y = revenue, fill = region)) +
  geom_boxplot() +
  labs(title = "Revenue by Region", x = "Region", y = "Revenue")
```

This helps me spot **outliers, distributions, and regional differences**.

## **4) Adding Interactivity with Plotly**

Static plots are good, but **interactive plots** let users explore deeper.

```{r}
library(plotly)

# Interactive scatter plot
p <- ggplot(df, aes(x = Marketing_Spend, y = revenue, color = region)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Marketing Spend vs Revenue")

ggplotly(p)

```
