---
title: "Survey Data - LikertData"
format: html
editor: visual
---

# **Analyzing Your Survey Data in R Programming: The Likert Package for Business Owners**

## Surveys are a vital tool for understanding customer sentiment. Here’s how to quantify that analysis in R programming

## You’ve sent out a customer satisfaction survey, an employee engagement questionnaire, or a product feedback form. The responses are now sitting in a Google Sheet or Excel file, and you’re staring at hundreds of rows of data, wondering how to make sense of it all. If your survey uses a rating scale — “Very Satisfied” to “Very Unsatisfied,” “Strongly Agree” to “Strongly Disagree” — you’re working with what data analysts call the Likert scale. The good news: R makes this remarkably straightforward.

## **Why Likert Scale Analysis Matters for Your Business**

Think of Likert analysis as a translator. Your raw survey responses tell an important story, but that story lives in scattered data points. A proper Likert analysis converts that noise into clear, visual insights you can share with stakeholders. Instead of saying “we have 847 survey responses,” you can say “72% of customers are satisfied or very satisfied, concentrated primarily in the Northeast region.” That specificity drives decisions.

Most business owners avoid analyzing survey data because the process seems technical. Between importing spreadsheets, transforming data formats, and wrestling with visualization tools, the effort-to-insight ratio feels unfavorable. The likert package in R eliminates this friction.

## **Getting Your Survey Data Into R**

If your survey lives in Google Sheets, you have two efficient paths. First, you can export your spreadsheet directly as a CSV file and import it into R using the simple *read.csv()* function. The second method is using the googlesheets4 package, which connects directly to your Google Sheet without downloading anything. This approach is particularly valuable if you have collected continuous feedback — the R analysis stays synchronized with your live data.

For Excel files, the readxl package handles import seamlessly. The syntax below shows an import of an example dataset:

```{r}
survey_data <- read_excel("customer_feedback.xlsx")

survey_data
```

## **Installing and Loading the Likert Package**

The installation follows in one of two ways — either via the Packages panel in RStudio (or similar for Positron) or via the install.packages() function. Here is the latter, calling the library on line two:

```{r}
# install.packages("likert")
library(likert)
library(readxl) # for reading data
library(dplyr) # for select and mutate functions
library(stringr) # trim whitespace
```

The *likert* package sits on top of ggplot2’s visualization foundation, inheriting the same clean, professional aesthetic that data-driven organizations expect. This package transforms raw survey response data into publication-ready visualizations and statistical summaries.

## **Preparing Your Survey Data**

The *likert* package expects your data in a specific format: each row represents a respondent, and each column represents a survey question. The responses should be factors (categorical data) with ordered levels — for example, “Strongly Disagree,” “Disagree,” “Neutral,” “Agree,” “Strongly Agree.”

```{r}
colnames(survey_data)
```

#### unique values

```{r}
unique(survey_data$question_1)
unique(survey_data$Region)
```

```{r}
# Define levels once to avoid typos
likert_levels <- c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree")

survey_clean <- survey_data |> 
  select(question_1, question_2, question_3, question_4) |> 
  mutate(across(everything(), 
                \(x) factor(x, levels = likert_levels, ordered = TRUE))) |>
  as.data.frame() # Important: likert() prefers a standard data.frame over a tibble

# Check levels before plotting
lapply(survey_clean, levels) 

likert_result <- likert(survey_clean)
plot(likert_result)

```

In the syntax below, I am telling R through the *likert()* function to use the Region column of the survey_data object as a Likert group. The 5 regions mentioned in the Region column represent where each respondent was located when the survey was taken. This information can help analysts and stakeholders gauge whether dissatisfaction or satisfaction was being driven by a particular region.

Below is the syntax and the resulting Likert plot for survey responses according to 5 regions.

```{r}
likert_by_region <- likert(survey_clean, grouping = survey_data$Region)
plot(likert_by_region)
```

## **Getting the Numbers Behind the Charts**

While visualizations communicate patterns, stakeholders often want the underlying statistics. To do so, use the *summary()* function to calculate percentages automatically. The image below shows the output from the *summary()* function for the *likert_result* object.

```{r}
summary(likert_result)
```

### Engagement

```{r}
#### Engagement

engagement <- read_excel("engagement_survey.xlsx")
colnames(engagement) <- c("Employee_ID", "leadership", "growth_opportunity", "work_environment", "compensation", "culture", "department", "tenure_years")
likert_levels <- c("Strongly Disagree", "Disagree", "Neutral", 
                   "Agree", "Strongly Agree")

engagement_clean <- engagement |>
  select(leadership, growth_opportunity, work_environment, 
         compensation, culture) |>
  mutate(across(everything(), 
                \(x) factor(stringr::str_trim(x), 
                            levels = likert_levels,
                            ordered = TRUE))) |>
  as.data.frame()  

# Analyze
engagement_likert <- likert(engagement_clean)

# Visualize
plot(engagement_likert)

```

```{r}
# Export summary
summary_stats <- summary(engagement_likert)
write.csv(summary_stats, "engagement_results.csv")
```
