---
title: "practical_packages"
format: html
editor: visual
---

## **The 3 Most Practical Packages for Data Visualization in R**

## **ggplot2 — The Foundation of Visualization**

`ggplot2` forms the basis of virtually all modern graphics in R. It processes data through a layer-by-layer logic ("grammar of graphics").

```{r}
library(ggplot2)
ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(title = "Engine Power and Fuel Efficiency",
       x = "Horsepower (hp)",
       y = "Miles Per Gallon (mpg)",
       color = "Number of Cylinders")
```

###  **plotly — Adding Interactivity**

Static graphs are excellent, but sometimes it is necessary to provide interactivity to the user. `plotly` transforms your ggplot2 graph into an interactive one with a single line:

```{r}
library(plotly)
p <- ggplot(mtcars, aes(hp, mpg, color = factor(cyl))) +
  geom_point(size = 3)
ggplotly(p)
```

The `echo: false` option disables the printing of code (only output is displayed).

### **ggiraph — Clickable Graphics**

`ggiraph` is ideal for web applications or Shiny projects. The points in your graph become clickable:

```{r}
# install.packages("ggiraph")
```

```{r}
library(ggiraph)
p <- ggplot(mtcars, aes(hp, mpg, tooltip = rownames(mtcars))) +
  geom_point_interactive(size = 3, color = "steelblue")
girafe(ggobj = p)
```
