---
title: "Raincloud Plots"
format: html
editor: visual
---

# **How to Create Raincloud Plots with ggrain for R Programming**

## Richer insights from data distributions are possible with this latest package for data analysis

Raincloud plots solve this kind of presentation challenge. They combine the quality of violin plots and boxplots to provide visualizations that address summary statistics and supporting insights.

The package carrying raincloud plot is a new one called *ggrain*. The *ggrain* package combines three essential visual elements — individual data points, distribution density, and summary statistics — into a single, comprehensive view. Instead of choosing between showing overall trends or individual details, you can present both simultaneously. For organizations making decisions based on performance distributions, customer behavior patterns, or quality control metrics, raincloud plots reveal the complete data story that summary statistics alone can’t tell.

## **Understanding Raincloud Plots**

Raincloud plots combine raw data points with two visualizations — violin plots that show data distribution and boxplots for displaying quartiles and outliers. The name raincloud comes from the visual appearance of the overall plot: the density curve resembles a cloud, while the individual points below look like falling rain.

Traditional visualizations force you to choose between showing aggregate patterns or individual observations. Bar charts show means but hide variability. Scatter plots show individual points but obscure overall trends. Boxplots reveal quartiles but can mask multimodal distributions where data clusters in multiple groups. Raincloud plots eliminate this trade-off by layering multiple visualization types.

The approach matters for business decisions because distributions often reveal operational realities that averages would otherwise obscure. A customer service team might show an average response time of 3 minutes, but a raincloud plot could reveal two distinct clusters: 80% of responses in under 2 minutes and 20% taking over 5 minutes. That bimodal distribution suggests different issues requiring different solutions — perhaps certain query types need specialized handling, or specific team members need additional training.

## **Installing and Loading ggrain**

The ggrain package requires ggplot2 as its foundation, building on the grammar of graphics framework for data visualization. So for a library liosting in a script, the *library()* function calls load both ggplot2 (required for the base plotting system) and ggrain (providing the specialized *geom_rain* function) into your R session.

```{r}
# install.packages("ggrain")
install.packages("remotes")
remotes::install_github("erocoar/gghalves")
```

libraries

```{r}
library(ggplot2)
library(ggrain)
```

## **Creating Your First Raincloud Plot**

The simplest raincloud plot requires just the *geom_rain* function added to your ggplot call. We’ll use the built-in iris dataset to demonstrate the basic syntax.

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_rain() +
  theme_minimal() +
  labs(
    title = "Sepal Width Distribution by Iris Species",
    x = "Species",
    y = "Sepal Width (cm)"
  )
```

The *ggplot()* function initializes the plot with the iris dataset and establishes aesthetic mappings through *aes()*. Setting *x = Species* creates categories along the horizontal axis, while *y = Sepal.Width* determines the vertical position of data points. The *fill = Species* aesthetic colors each species distinctly, improving visual separation. The *geom_rain()* function generates all three visualization components—violin plot, boxplot, and individual points—automatically.

The function handles positioning, scaling, and layering without requiring manual adjustments, making it straightforward to create publication-quality visualizations. The *theme_minimal()* reduces visual clutter by removing unnecessary grid lines, and *labs()* adds clear axis labels and titles that make the plot self-explanatory for stakeholders.

## **Customizing Individual Components**

Each element of a raincloud plot can be customized independently through specialized argument lists. This granular control lets you emphasize specific aspects of your data story.

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_rain(
    alpha = 0.5,
    violin.args = list(alpha = 0.3, color = NA),
    boxplot.args = list(outlier.shape = NA),
    point.args = list(size = 1.5, alpha = 0.6)
  ) +
  theme_minimal() +
  labs(
    title = "Sepal Width Distribution by Iris Species",
    x = "Species",
    y = "Sepal Width (cm)"
  )
```

The *violin.args* parameter accepts a list of modifications for the density curve component. Setting *alpha = 0.3* makes the violin semi-transparent, preventing it from obscuring underlying points, while *color = NA* removes the outline for a cleaner appearance. The *boxplot.args* parameter controls the boxplot layer—here, *outlier.shape = NA* prevents double-plotting of outliers that already appear as individual points. The *point.args* parameter adjusts the raw data points, where *size = 0.5* makes them more visible and *alpha = 0.6* adds transparency to reveal overlapping observations. This layered customization approach gives you precise control over visual hierarchy, letting you emphasize the elements most relevant to your analysis while maintaining the comprehensive view that makes raincloud plots valuable.

## **Adding a Covariate for Color Coding**

When your analysis involves a third variable, ggrain can color individual points based on that covariate while maintaining group separation.

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_rain(cov = "Sepal.Length") +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(
    title = "Sepal Width Distribution by Iris Species",
    x = "Species",
    y = "Sepal Width (cm)"
  )
```

The *cov* parameter specifies which variable should determine point colors. By setting *cov = "Sepal.Length"* , each point's color reflects its sepal length value, creating a gradient that reveals relationships between variables within each species group. The *scale_color_viridis_c()* function applies a perceptually-uniform color scale that remains interpretable for colorblind viewers and in grayscale printing. This approach is particularly valuable for business applications where you need to show how a secondary metric varies within performance groups—for example, showing how deal size (color) varies within sales rep performance tiers (x-axis groupings), or how customer satisfaction scores (color) distribute within different product categories (groups).

## **Controlling Rain Direction and Position**

Raincloud orientation affects how easily viewers can compare distributions across groups. The *rain.side* parameter provides flexible positioning options.

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_rain(rain.side = "l") +
  theme_minimal() +
  labs(
    title = "Sepal Width Distribution by Iris Species",
    x = "Species",
    y = "Sepal Width (cm)"
  )
```

The *rain.side* parameter accepts four values that control component positioning: "r" (right, the default) places density curves to the right of center, "l" (left) mirrors them to the left, "f" (flanking) creates symmetric raincloud plots with distributions on both sides, and "f1x1" or "f2x2" create specialized paired comparisons. Left-facing plots can improve readability when you have many groups, as the density curves don't overlap with axis labels on the right. Flanking configurations work well for before-after comparisons or paired data, showing both distributions simultaneously for direct visual comparison without requiring separate plot panels.

## **Working with Paired Data**

Paired observations — like before-and-after measurements or test-retest data — require special handling to show the relationships between matched observations.

```{r}
ggplot(sleep, aes(x = group, y = extra, fill = group)) +
  geom_rain(id.long.var = "ID") +
  theme_minimal() +
  labs(
    title = "Sleep Study: Hours of Extra Sleep by Drug Group",
    subtitle = "Lines connect the same individual and across conditions",
    x = "Drug Group",
    y = "Extra Hours of Sleep"
  )
```

The *id.long.var* parameter identifies which variable contains the pairing information, allowing ggrain to draw connecting lines between matched observations. Setting *id.long.var = ID* tells the function to connect points sharing the same ID value, making individual trajectories visible within the aggregate distribution. These connecting lines reveal whether most individuals follow the group trend or if the aggregate pattern masks varied individual responses. For business applications, this matters when analyzing A/B test results where the same customers experience both conditions, or when tracking employee performance before and after training interventions. The paired visualization reveals whether changes are consistent across individuals or driven by specific subgroups.

## **Adjusting Positioning and Width**

Fine-tuning the spatial arrangement of raincloud components helps prevent overcrowding and improves visual clarity when working with multiple groups.

```{r}
ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_rain(
    boxplot.args.pos = list(width = 0.1, position = position_nudge(x = -0.2)),
    violin.args.pos = list(side = "r", width = 0.7, position = position_nudge(x = 0.15))
  ) +
  theme_minimal() +
  labs(
    title = "Petal Length with Adjusted Component Spacing",
    x = "Species",
    y = "Petal Length (cm)"
  )
```

The code creates the following plot. The *boxplot.args.pos* parameter controls boxplot positioning through two mechanisms: *width* determines the boxplot's horizontal extent (0.1 creates narrow boxes that don't dominate the plot), while *position_nudge(x = -0.2)* shifts the entire boxplot left by 0.2 units to prevent overlap with points.

## **Handling Categorical Data with Likert Scales**

Survey data and Likert-scale responses create unique visualization challenges because values are discrete rather than continuous. The ggrain package includes specific handling for these data types.

```{r}
survey_data <- data.frame(
  response = sample(1:5, 200, replace = TRUE, prob = c(0.1, 0.2, 0.4, 0.2, 0.1)),
  department = rep(c("Sales", "Marketing", "Operations", "IT"), each = 50)
)

ggplot(survey_data, aes(x = department, y = response, fill = department)) +
  geom_rain(likert = TRUE) +
  scale_y_continuous(breaks = 1:5, labels = c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strondly Agree"))

theme_minimal() +
  labs(
    title = "Employee Satisfaction by Department",
    x = "Department",
    y = "Satisfaction level"
  )
```

The *likert = TRUE* parameter activates specialized handling for discrete, ordered categorical data. This setting adds vertical jitter to points that would otherwise stack directly on top of each other at integer values, making individual responses visible while maintaining their categorical nature. The jittering is constrained to prevent points from appearing at impossible values (like 2.5 on a 1-5 scale), preserving the interpretability of the ordinal scale. The *scale_y_continous()* function replaces numeric axis labels with meaningful category names, making the plot immediately understandable for non-technical stakeholders. For organizations analyzing employee surveys, customer satisfaction scores, or Net Promoter Score distributions, this feature transforms unusable stacked-point charts into informative visualizations that reveal both group tendencies and response variability.

## **Integrating with ggplot2 Themes and Scales**

Raincloud plots integrate seamlessly with ggplot2’s theming system and scale modifications, maintaining consistency with your organization’s visual standards.

```{r}
ggplot(iris, aes(x = Species, y = Petal.Width, fill = Species)) +
  geom_rain(
    alpha = 0.6,
    violin.args = list(alpha = 0.4, color = "white", linewidth = 0.5),
    point.args = list(size = 1, alpha = 0.7)
  ) +
  scale_fill_manual(values = c("#2e86ab", "#a23b72", "#f18f01")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Pedal Width Distribition across Iris Species",
    subtitle = "Raincloud visualization showing distribution, quartiles, and individual measurements",
    x = "Species",
    y = "Petal Width (cm)",
    caption = "Data source: Fisher's Iris dataset"
  )
```

The s*cale_fill_manual()* function applies custom colors that match corporate branding guidelines, replacing ggplot2's default color palette with specific hex values. The *theme()* function modifies specific plot elements: *plot.title* increases title size and adds bold formatting for emphasis, axis.title ensures readable axis labels, *legend.position = "bottom"* moves the legend to prevent obscuring data, and *panel.gird.minor = element_blank()* removes minor grid lines that add clutter without improving interpretability. The *labs()* function adds a subtitle explaining what viewers see and a caption documenting data sources, which matters for reproducibility and establishing credibility with stakeholders. This comprehensive styling approach ensures your raincloud plots maintain visual consistency with other reports while meeting publication standards for presentations, white papers, or research documents.

## **Comparing Multiple Groups with Faceting**

When you need to compare distributions across multiple categorical dimensions simultaneously, faceting creates small multiples that maintain visual clarity.

The code below creates a sample data frame of revenue for each region, then creates a visualization displaying each group via faceting.

```{r}
set.seed(123)
performance_data <- data.frame(
  region = rep(c("North", "South", "East", "West"), each = 75),
  quarter = rep(rep(c("Q1", "Q2", "Q3"), each = 25), 4),
  revenue = c(
    rnorm(75, mean = 50000, sd = 8000), # North
    rnorm(75, mean = 45000, sd = 7000), # south
    rnorm(75, mean = 55000, sd = 9000), # east
    rnorm(75, mean = 48000, sd = 7500) # west
  )
)

ggplot(performance_data, aes(x = quarter, y = revenue, fill = quarter)) +
  geom_rain(alpha = 0.5) +
  facet_wrap(~ region, ncol = 2) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size =11),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Quarterly Revenue Distribution by Region",
    x = "Quarter",
    y = "Revenue",
    fill = "Quarter"
  )
```

The *facet_wrap()* function creates separate panels for each region, using `~ region` to specify the faceting variable and *ncol = 2* to arrange panels in a two-column grid. This layout enables direct comparison of quarterly patterns across regions without cramming all distributions into a single crowded plot. The *scales::dollar_format()* function formats y-axis labels as currency, improving readability for business stakeholders who think in dollar amounts rather than scientific notation. The *strip.text* theme modification increases facet label size and adds bold formatting so readers can easily identify which panel shows which region. This multi-panel approach scales well when you need to compare patterns across time periods, product lines, customer segments, or geographic regions—common requirements in business analytics where understanding variation across multiple dimensions drives strategic decisions.

## **Practical Applications for Business Analysis**

Raincloud plots solve specific visualization challenges across multiple business domains. Understanding when they provide unique value helps you choose the right tool for your analytical needs.

**Quality Control and Manufacturing**: When monitoring product specifications, raincloud plots reveal whether variation stays within acceptable limits and whether the distribution shifts over time. A manufacturing process might show acceptable average dimensions but hide a bimodal distribution indicating two distinct machine settings that need alignment.

**Sales Performance Analysis**: Traditional bar charts showing average sales per rep obscure the distribution of deal sizes and rep consistency. Raincloud plots reveal whether high average performance comes from consistent mid-level deals or from landing occasional large contracts interspersed with dry periods.

**Customer Experience Metrics**: Survey responses often show ceiling effects where most scores cluster at the top but a small group of detractors drives down the average. Raincloud plots make these subgroups visible, helping you understand whether you need broad improvements or targeted interventions for specific customer segments.

**Clinical and Health Data**: Medical research frequently involves comparing treatment effects where individual variation matters as much as group differences. Raincloud plots show whether a treatment produces consistent benefits across patients or helps only specific subgroups.

**A/B Testing Results**: When comparing conversion rates or user engagement metrics across test variants, raincloud plots reveal whether observed differences represent genuine effects or result from a few extreme observations that skew the means.
