---
title: "Seasonal Forecasting"
format: html
editor: visual
---

## **How to Use Triple Exponential Smoothing for Seasonal Forecasting in R Programming**

#### Smoothing time series data is an essential task for determining a trend in a dataset. Learn how to use smoothing as a forecast technique.

## **Implementing HoltWinters() from the Forecast Library**

The forecast library provides the foundational *HoltWinters()* function for triple exponential smoothing in R. This function requires a time series object created with the `ts()` function, where you specify the frequency parameter matching your seasonal cycle—12 for monthly data with yearly seasonality, 4 for quarterly data, or 7 for daily data with weekly patterns.

```{r}
# install.packages("forecast")
```

first code

```{r}
library(forecast)

# create time series object with monthly data
sales_data <- c(142, 158, 165, 178, 189, 195,
                198, 205, 220, 245, 280, 310,
                155, 172, 180, 192, 205, 210,
                215, 222, 235, 260, 295, 325)

sales_ts <- ts(sales_data, start = c(2023, 1), frequency = 12)

# Fit Holt-Winters model
hw_model <- HoltWinters(sales_ts)

# Generate forecasts for next 6 months
hw_forecast <- forecast(hw_model, h = 6)

# View model summary and forecast
summary(hw_model)
print(hw_forecast)

# Plot forecasts with confidence intervals
plot(hw_forecast, main = "Sales Forecast with Triple Exponential Smoothing")
```

The `echo: false` option disables the printing of code (only output is displayed).
