---
title: "background maps"
format: html
editor: visual
---

## Background maps

# Most basic background map with R and Leaflet

This post is a very basic introduction to the `leaflet` R package. It shows how to display a [background map](https://r-graph-gallery.com/map.html) using default parameters, with reproducible code provided.

This page explains how to display a [background map](https://r-graph-gallery.com/map.html) with the R `leaflet` package.

First initiate the map with the `leaflet()` function. Then add tiles with `addTiles()`. Note the use of the `%>%` operator to “pipe” functions. Here you can see how to get the same result using or not this operator.

By default, you get the map beside. See next charts to learn how to [zoom on a zone](https://r-graph-gallery.com/179-2-select-a-zone-in-leaflet-map.html), change [background style](https://r-graph-gallery.com/180-change-background-in-leaflet-map.html).

```{r}
# Load the library
library(leaflet)
 
# Note: if you do not already installed it, install it with:
# install.packages("leaflet")

# Initialize the leaflet map with the leaflet() function
m <- leaflet()
# Then we Add default OpenStreetMap map tiles
m <- addTiles(m)
m
 
# Same stuff but using the %>% operator
m <- leaflet() %>% 
  addTiles()
m

# save the widget in a html file if needed.
# library(htmlwidgets)
# saveWidget(m, file=paste0( getwd(), "/HtmlWidget/backgroundMapBasic.html"))
```

### Example for Costa Rica

Zoom levels in Leaflet typically range from `0` (the whole world) to `19` (a single building). A zoom level of **`7`** or **`8`** is ideal for framing the entire country of Costa Rica beautifully in the viewer.

```{r}
library(leaflet)

# Initialize the leaflet map and set the view to Costa Rica
m <- leaflet() %>% 
  addTiles() %>% 
  # lng = Longitude, lat = Latitude, zoom = 7 is perfect for a small country
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

# Print the map
m

# Optional: Save the widget to your working directory
# library(htmlwidgets)
# saveWidget(m, file = "costa_rica_map.html")
```

### Costa Rica with black and white tile

```{r}
library(leaflet)

# Choice 1: A beautiful, clean light-grey/minimalist style
m <- leaflet() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 8)

m

# Choice 2: If you want a literal high-contrast Black & White map
m_bw <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_bw
```

### City view

```{r}
m_sf10 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 10)

m_sf10

m_sf13 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 13)

m_sf13

m_sf16 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 16)

m_sf16

m_sf19 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 19)

m_sf19

```

```{r}
m_sf13 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = 37.76647371951237, lat = -122.44198850042363, zoom = 13)

m_sf13
```
