---
title: "background maps"
format: html
editor: visual
---

## Background maps

# Most basic background map with R and Leaflet

This post is a very basic introduction to the `leaflet` R package. It shows how to display a [background map](https://r-graph-gallery.com/map.html) using default parameters, with reproducible code provided.

This page explains how to display a [background map](https://r-graph-gallery.com/map.html) with the R `leaflet` package.

First initiate the map with the `leaflet()` function. Then add tiles with `addTiles()`. Note the use of the `%>%` operator to “pipe” functions. Here you can see how to get the same result using or not this operator.

By default, you get the map beside. See next charts to learn how to [zoom on a zone](https://r-graph-gallery.com/179-2-select-a-zone-in-leaflet-map.html), change [background style](https://r-graph-gallery.com/180-change-background-in-leaflet-map.html).

```{r}
# Load the library
library(leaflet)
 
# Note: if you do not already installed it, install it with:
# install.packages("leaflet")

# Initialize the leaflet map with the leaflet() function
m <- leaflet()
# Then we Add default OpenStreetMap map tiles
m <- addTiles(m)
m
 
# Same stuff but using the %>% operator
m <- leaflet() %>% 
  addTiles()
m

# save the widget in a html file if needed.
# library(htmlwidgets)
# saveWidget(m, file=paste0( getwd(), "/HtmlWidget/backgroundMapBasic.html"))
```

### Example for Costa Rica

Zoom levels in Leaflet typically range from `0` (the whole world) to `19` (a single building). A zoom level of **`7`** or **`8`** is ideal for framing the entire country of Costa Rica beautifully in the viewer.

```{r}
library(leaflet)

# Initialize the leaflet map and set the view to Costa Rica
m <- leaflet() %>% 
  addTiles() %>% 
  # lng = Longitude, lat = Latitude, zoom = 7 is perfect for a small country
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

# Print the map
m

# Optional: Save the widget to your working directory
# library(htmlwidgets)
# saveWidget(m, file = "costa_rica_map.html")
```

### Costa Rica with black and white tile

```{r}
library(leaflet)

# Choice 1: A beautiful, clean light-grey/minimalist style
m <- leaflet() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 8)

m

# Choice 2: If you want a literal high-contrast Black & White map
m_bw <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_bw
```

### City view

```{r}
m_sf10 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 10)

m_sf10

m_sf13 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 13)

m_sf13

m_sf16 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 16)

m_sf16

m_sf19 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 19)

m_sf19

```

```{r}
m_sf13 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = 37.76647371951237, lat = -122.44198850042363, zoom = 13)

m_sf13
```

# Change and customize circles and rectangles on a leaflet map

Learn how to change and customize circles and rectangles on a `leaflet` map. Three examples are provided: **basic** circle, **custom** circle and circle with **size that change** when you zoom.

# List of tiles

Let’s start with basic circle (default propose by `leaflet`). Please note that since we use here the `addCircleMarkers()` function, the **size of the circle** are set in pixels. It means that markers will keep always the **same size**, even when you zoom in or out!

See below in order to find how to change the size of the circle, **dynamically or not**.

```{r}
#library
library(leaflet)
 
# Create 20 markers (Random points)
data = data.frame(
   long=sample(seq(-150,150),20),
   lat=sample(seq(-50,50),20),
   val=round(rnorm(20),2),
   name=paste("point",letters[1:20],sep="_")
) 
 
# Show a circle at each position
m = leaflet(data = data) %>%
   addTiles() %>%
   addCircleMarkers(~long, ~lat , popup = ~as.character(name))
m

# save the widget
# library(htmlwidgets)
# saveWidget(m, file="HtmlWidget/leaflet-182-add-circles.html", selfcontained = F)
```

# Custom circles

It is possible to custom the circles. Really helpful since it permits to change it **following the value of a variable** from the dataset. Here, the size of the circle is defined by the `val`

```{r}
# Create 20 markers (Random points)
data = data.frame(
   long=sample(seq(-150,150),20),
   lat=sample(seq(-50,50),20),
   val=round(rnorm(20),2),
   name=paste("point",letters[1:20],sep="_")
) 
 
# Show a CUSTOM circle at each position. Size defined in Pixel. Size does not change when you zoom
m=leaflet(data = data) %>%
   addTiles() %>%
   addCircleMarkers(
      ~long, ~lat, 
      radius=~val*14 , 
      color=~ifelse(data$val>0 , "red", "orange"),
      stroke = TRUE, 
      fillOpacity = 0.2,
      popup = ~as.character(name)
   ) 
m

# save the widget
# library(htmlwidgets)
# saveWidget(m, file="HtmlWidget/leaflet-182-add-circles-2.html", selfcontained = F)
```

# Custom circles with size that change when you zoom

You can also add circles with the `addCircle()` function. This time, the radius of the circle is **defined in meters**. It means that its **size will change** when you zoom on the map!

```{r}
#library
library(leaflet)
 
# Create 20 markers (Random points)
data = data.frame(
   long=sample(seq(-150,150),20),
   lat=sample(seq(-50,50),20),
   val=round(rnorm(20),2),
   name=paste("point",letters[1:20],sep="_")
) 
 
# Show a CUSTOM circle at each position. Size in meters --> change when you zoom.
m = leaflet(data = data) %>%
   addTiles() %>%
   addCircles(
      ~long, ~lat, 
      radius=~val*1000000 , 
      color=~ifelse(data$val>0 , "red", "orange"),
      stroke = TRUE, 
      fillOpacity = 0.2,
      popup = ~as.character(name)
   ) %>% 
  setView(lng = 166.45, lat = 21, zoom = 2)
m

# save the widget
# library(htmlwidgets)
# saveWidget(m, file="HtmlWidget/leaflet-182-add-circles-3.html", selfcontained = F)
```

# Add rectangles

Lastly, how to **add a rectangle** with the `addRectangles()` function. You have to give the **2 coordinates of 2 opposites** corners of the rectangle.

```{r}
#library
library(leaflet)
 
# Create 20 markers (Random points)
data = data.frame(
   long=sample(seq(-150,150),20),
   lat=sample(seq(-50,50),20),
   val=round(rnorm(20),2),
   name=paste("point",letters[1:20],sep="_")
) 
 
# Show a rectangle
m = leaflet() %>% addTiles() %>%  
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  addRectangles(
    lng1=-118.456554, lat1=34.078039,
    lng2=-118.436383, lat2=34.062717,
    fillColor = "transparent"
  )
m

# save the widget
# library(htmlwidgets)
# saveWidget(m, file="HtmlWidget/leaflet-182-add-rectangles.html", selfcontained = F)
```

# Loading a tile

This post follows the previous posts on `leaflet`: most [basic map](https://r-graph-gallery.com/179-show-a-map-with-leaflet-r.html), and zooming on a [specific area](https://r-graph-gallery.com/179-2-select-a-zone-in-leaflet-map.html).

`Leaflet` offers several tiles to customize the background used on a map.

A tile can be loaded thanks to the `addProviderTiles()` function.

```{r}
# Load the library
library(leaflet)

# Note: if you do not already installed it, install it with:
# install.packages("leaflet")

# Background 1: NASA
m <- leaflet() %>% 
   addTiles() %>% 
   setView( lng = 2.34, lat = 48.85, zoom = 5 ) %>% 
   addProviderTiles("NASAGIBS.ViirsEarthAtNight2012")
m
 
# Background 2: World Imagery
m <- leaflet() %>% 
   addTiles() %>% 
   setView( lng = 2.34, lat = 48.85, zoom = 3 ) %>% 
   addProviderTiles("Esri.WorldImagery")
m

# costa rica
m_cr <- leaflet() %>% 
  addProviderTiles("NASAGIBS.ViirsEarthAtNight2012") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_sf10 <- leaflet() %>% 
  addProviderTiles("NASAGIBS.ViirsEarthAtNight2012") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_sf10

# save the widget in a html file if needed.
# library(htmlwidgets)
# saveWidget(m, file=paste0( getwd(), "/HtmlWidget/backgroundMapTile.html", width="1000px"))
```

# List of tiles

Many tiles are made available. A complete list can be found [here](http://leaflet-extras.github.io/leaflet-providers/preview/index.html). Most famous are probably:

- Nasa: `NASAGIBS.ViirsEarthAtNight2012`

- Google map: `Esri.WorldImagery`

- Gray: `Esri.WorldGrayCanvas`

- Terrain: `Esri.WorldTerrain`

- Topo Map: `Esri.WorldTopoMap`

```{r}
# costa rica
m_cr <- leaflet() %>% 
  addProviderTiles("NASAGIBS.ViirsEarthAtNight2012") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldGrayCanvas") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldTerrain") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldTopoMap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerBlacklite") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerDark") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Stadia.StamenWatercolor") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Thunderforest.OpenCycleMap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Thunderforest.SpinalMap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Thunderforest.MobileAtlas") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Thunderforest.Neighbourhood") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("CyclOSM") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr
```

more tiles

```{r}
m_cr <- leaflet() %>% 
  addProviderTiles("Jawg.Matrix") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldStreetMap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.OceanBasemap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("CartoDB.PositronNoLabels") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("CartoDB.DarkMatter") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr
```

more

```{r}
m_cr <- leaflet() %>% 
  addProviderTiles("OpenRailwayMap") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("WaymarkedTrails.skating") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("OpenRailwayMap") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("SafeCast") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("WaymarkedTrails.cycling") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("WaymarkedTrails.hiking") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr
```

# The `Maps`, `Mapdata` and `Oz` packages to get most common boundaries

The [maps](https://r-graph-gallery.com/278-the-maps-library.html#maps), [mapdata](https://r-graph-gallery.com/278-the-maps-library.html#mapdata) and [oz](https://r-graph-gallery.com/278-the-maps-library.html#oz) packages provide the boundaries of the most common world regions like the US, Europe, New Zealand and more. This post describes how to use them to avoid struggling finding input data.

# `maps` package

The [maps](https://cran.r-project.org/web/packages/maps/maps.pdf) package is the best source of geospatial data in R.

The whole list of offered data is available typing: `help(package='maps')`.

It includes:

- World: `world`, `world.cities`, `lakes`

- US: states, `county`, `state`, `usa`

- France: `france`

- Italy: `italy`

- New zealand: `nz`

The code below shows how to use one of this geospatial object.

```{r}
help(package='maps')
```

```{r}
# Load library
library(maps)

# Check all available geospatial objects:
# help(package='maps')

# Map of the world:
map('world',col="grey", fill=TRUE, bg="white", lwd=0.05, mar=rep(0,4),border=0, ylim=c(-80,80) )
```

# `mapdata` package

The [mapdata](https://cran.r-project.org/web/packages/mapdata/mapdata.pdf) package extends the `maps` package with more geospatial datasets:

- `china`

- `japan`

- Other world versions like pacific Centric (`world2Hires`)

See complete list with `help(package='mapdata')`

```{r}
help(package='mapdata')
```

```{r}
# Load library
library(mapdata)

# Check all available geospatial objects:
# help(package='mapdata')

# Map of Japan:
map('japan',col="black", lwd=1, mar=rep(0,4) )
```

```{r}
# Load library
library(mapdata)

# Check all available geospatial objects:
# help(package='mapdata')

# Map of Japan:
map('china',col="black", lwd=1, mar=rep(0,4) )
```

# `oz` package

The [oz](https://github.com/cran/oz) package provides several geospatial object concerning Australia.

Plot the general country using the `oz()` function, or states using `nsw()`, `qld()` and so on..

Type `help(package='oz')` for documentation.

```{r}
install.packages('oz')
```

```{r}
help(package='oz')
```

```{r}
# Load library
library(oz)

# Check all available geospatial objects:
# help(package='oz')

# Map of Australia
par(mar=rep(0,4))
oz( states=TRUE, col="#69b3a2")
```

# Open and Plot Geojson files in R

Geojson files are a common way to store geospatial data. This post explains how to read it with `R` and the `sf` package, and how to plot it in base R or with `ggplot2`.

If you did not find the geospatial data you need in existing R packages (see the [map section](https://r-graph-gallery.com/map.html)), you need to find this information elsewhere on the web.

It will often be stored as a `.geoJSON` format. This post explains how to read it.

# Find and download a `.geoJSON` file

You need to dig the internet to find the geoJSON file you are interested in. For instance, [this URL](https://raw.githubusercontent.com/gregoiredavid/france-geojson/master/communes.geojson) provides a file containing french region boundaries.

You can load it in R with:

```{r}
# Download to a temporary file
tmp_geojson <- tempfile(fileext = ".geojson")
download.file(
  "https://raw.githubusercontent.com/gregoiredavid/france-geojson/master/communes.geojson",
  tmp_geojson
)

# Let's read the downloaded geoJson file with the sf library:
library(sf)
my_sf <- read_sf(tmp_geojson)
my_sf
```

```{r}
# View(my_sf)

# Force the console to print wider rows
options(width = 1000)

# Print the first few rows
head(as.data.frame(my_sf))

library(tibble)

# This prints every column vertically, showing the full content of the first few rows
glimpse(my_sf)
```

That’s it! You now have a geospatial object called `my_sf`. I strongly advise to read [this post](https://r-graph-gallery.com/169-170-basic-manipulation-of-shapefiles) to learn how to manipulate it.

Just in case, here is how to plot it in base R and with `ggplot2`.

# Plot it with `base R`

The basic `plot()` function knows how to plot a geospatial object. Thus you just need to pass it `my_sf` and add a couple of options to customize the output.

```{r}
# Select only the region #6 as you would do with a data frame
my_sf_region_6 <- my_sf[substr(my_sf$code, 1, 2) == "10", ]

par(mar = c(0, 0, 0, 0))
plot(my_sf_region_6$geometry, col = "#9CB6D9")
```

It is totally possible (and advised IMO) to build the map with [ggplot2](https://r-graph-gallery.com/ggplot2-package.html), using the `geom_sf()` function as described below.

```{r}
# Plot it
library(ggplot2)
ggplot(my_sf_region_6) +
  geom_sf(fill = "#69b3a2", color = "white") +
  theme_void()
```

# Open and Plot Shapefiles in R

**Shapefiles**, a popular format for storing **geospatial information**, are easily handled in **R** using the `sf` package. This guide will demonstrate **how to read shapefiles** into R and visualize them using either **base R** or the [ggplot2](https://r-graph-gallery.com/ggplot2-package.html) package.

# About shapefiles

------------------------------------------------------------------------

If you did not find the geospatial data you need in existing R packages (see the [map section](https://r-graph-gallery.com/map.html)), you need to find this information elsewhere on the web.

Usually, you will find it as a *shape file* format. This format is composed by **several files** that you need to **keep together** in the same folder.

# Find and download a shapefile.

------------------------------------------------------------------------

You need to dig the internet to find the shape file you are interested in. For instance, [this URL](http://thematicmapping.org/downloads/world_borders.php) will redirect you to a zipped shape file containing the world boundaries.

more maps - <https://mapcruzin.com/download-mexico-canada-us-transportaton-shapefile.htm#mex>

You can download it and unzip it with R:

```{r}
# Download the shapefile. (note that I store it in a folder called DATA. You have to change that if needed.)
download.file("https://mapcruzin.com/download-mexico-canada-us-shapefile/mexico-states.zip", destfile = "DATA/world_shape_file.zip")
# You now have it in your current working directory, have a look!

# Unzip this file. You can do it with R (as below), or clicking on the object you downloaded.
unzip("DATA/world_shape_file.zip", junkpaths = FALSE)
#  -- > You now have 4 files. One of these files is a .shp file! (TM_WORLD_BORDERS_SIMPL-0.3.shp)
```

# Read it with `sf`

------------------------------------------------------------------------

The `sf` package offers the `st_read()` and the `read_sf()` functions that allows to **read shapefile** using the following syntax.

As a result you get a geospatial object (`my_sf` here) that contains all the information we need for further mapping. Please try the following command to understand how this object works:

- `print(my_sf)`: tells you the max and min coordinates, the kind of projection in use and the the **first few rows** of the data associated with the regions.

- `nrow(my_sf)`: how many regions you have

```{r}
# Read this shape file with the sf library.
library(sf)
my_sf <- read_sf("mexbdy.shp")

# -- > Now you have a sf object (simple feature data frame). You can start doing maps!

my_sf
```

# Plot it with `base R`

------------------------------------------------------------------------

The basic `plot()` function knows how to plot a geospatial object. Thus you just need to pass it `my_sf` and add a couple of options to customize the output.

```{r}
# Basic plot of this shape file:
par(mar = c(0, 0, 0, 0))
plot(st_geometry(my_sf), col = "#f2f2f2", bg = "skyblue", lwd = 0.25, border = 0)
```

# Plot it with `ggplot2`

It is totally possible (and advised) to build the map with [ggplot2](https://r-graph-gallery.com/ggplot2-package.html), using the `geom_sf()` function as described below.

```{r}
# Plot it
library(ggplot2)
ggplot(my_sf) +
  geom_sf(fill = "#69b3a2", color = "white") +
  theme_void()
```
