---
title: "background maps"
format: html
editor: visual
---

## Background maps

# Most basic background map with R and Leaflet

This post is a very basic introduction to the `leaflet` R package. It shows how to display a [background map](https://r-graph-gallery.com/map.html) using default parameters, with reproducible code provided.

This page explains how to display a [background map](https://r-graph-gallery.com/map.html) with the R `leaflet` package.

First initiate the map with the `leaflet()` function. Then add tiles with `addTiles()`. Note the use of the `%>%` operator to “pipe” functions. Here you can see how to get the same result using or not this operator.

By default, you get the map beside. See next charts to learn how to [zoom on a zone](https://r-graph-gallery.com/179-2-select-a-zone-in-leaflet-map.html), change [background style](https://r-graph-gallery.com/180-change-background-in-leaflet-map.html).

```{r}
# Load the library
library(leaflet)
 
# Note: if you do not already installed it, install it with:
# install.packages("leaflet")

# Initialize the leaflet map with the leaflet() function
m <- leaflet()
# Then we Add default OpenStreetMap map tiles
m <- addTiles(m)
m
 
# Same stuff but using the %>% operator
m <- leaflet() %>% 
  addTiles()
m

# save the widget in a html file if needed.
# library(htmlwidgets)
# saveWidget(m, file=paste0( getwd(), "/HtmlWidget/backgroundMapBasic.html"))
```

### Example for Costa Rica

Zoom levels in Leaflet typically range from `0` (the whole world) to `19` (a single building). A zoom level of **`7`** or **`8`** is ideal for framing the entire country of Costa Rica beautifully in the viewer.

```{r}
library(leaflet)

# Initialize the leaflet map and set the view to Costa Rica
m <- leaflet() %>% 
  addTiles() %>% 
  # lng = Longitude, lat = Latitude, zoom = 7 is perfect for a small country
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

# Print the map
m

# Optional: Save the widget to your working directory
# library(htmlwidgets)
# saveWidget(m, file = "costa_rica_map.html")
```

### Costa Rica with black and white tile

```{r}
library(leaflet)

# Choice 1: A beautiful, clean light-grey/minimalist style
m <- leaflet() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 8)

m

# Choice 2: If you want a literal high-contrast Black & White map
m_bw <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_bw
```

### City view

```{r}
m_sf10 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 10)

m_sf10

m_sf13 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 13)

m_sf13

m_sf16 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 16)

m_sf16

m_sf19 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 19)

m_sf19

```

```{r}
m_sf13 <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  setView(lng = 37.76647371951237, lat = -122.44198850042363, zoom = 13)

m_sf13
```

# Change and customize circles and rectangles on a leaflet map

Learn how to change and customize circles and rectangles on a `leaflet` map. Three examples are provided: **basic** circle, **custom** circle and circle with **size that change** when you zoom.

# List of tiles

Let’s start with basic circle (default propose by `leaflet`). Please note that since we use here the `addCircleMarkers()` function, the **size of the circle** are set in pixels. It means that markers will keep always the **same size**, even when you zoom in or out!

See below in order to find how to change the size of the circle, **dynamically or not**.

```{r}
#library
library(leaflet)
 
# Create 20 markers (Random points)
data = data.frame(
   long=sample(seq(-150,150),20),
   lat=sample(seq(-50,50),20),
   val=round(rnorm(20),2),
   name=paste("point",letters[1:20],sep="_")
) 
 
# Show a circle at each position
m = leaflet(data = data) %>%
   addTiles() %>%
   addCircleMarkers(~long, ~lat , popup = ~as.character(name))
m

# save the widget
# library(htmlwidgets)
# saveWidget(m, file="HtmlWidget/leaflet-182-add-circles.html", selfcontained = F)
```

# Custom circles

It is possible to custom the circles. Really helpful since it permits to change it **following the value of a variable** from the dataset. Here, the size of the circle is defined by the `val`

```{r}
# Create 20 markers (Random points)
data = data.frame(
   long=sample(seq(-150,150),20),
   lat=sample(seq(-50,50),20),
   val=round(rnorm(20),2),
   name=paste("point",letters[1:20],sep="_")
) 
 
# Show a CUSTOM circle at each position. Size defined in Pixel. Size does not change when you zoom
m=leaflet(data = data) %>%
   addTiles() %>%
   addCircleMarkers(
      ~long, ~lat, 
      radius=~val*14 , 
      color=~ifelse(data$val>0 , "red", "orange"),
      stroke = TRUE, 
      fillOpacity = 0.2,
      popup = ~as.character(name)
   ) 
m

# save the widget
# library(htmlwidgets)
# saveWidget(m, file="HtmlWidget/leaflet-182-add-circles-2.html", selfcontained = F)
```

# Custom circles with size that change when you zoom

You can also add circles with the `addCircle()` function. This time, the radius of the circle is **defined in meters**. It means that its **size will change** when you zoom on the map!

```{r}
#library
library(leaflet)
 
# Create 20 markers (Random points)
data = data.frame(
   long=sample(seq(-150,150),20),
   lat=sample(seq(-50,50),20),
   val=round(rnorm(20),2),
   name=paste("point",letters[1:20],sep="_")
) 
 
# Show a CUSTOM circle at each position. Size in meters --> change when you zoom.
m = leaflet(data = data) %>%
   addTiles() %>%
   addCircles(
      ~long, ~lat, 
      radius=~val*1000000 , 
      color=~ifelse(data$val>0 , "red", "orange"),
      stroke = TRUE, 
      fillOpacity = 0.2,
      popup = ~as.character(name)
   ) %>% 
  setView(lng = 166.45, lat = 21, zoom = 2)
m

# save the widget
# library(htmlwidgets)
# saveWidget(m, file="HtmlWidget/leaflet-182-add-circles-3.html", selfcontained = F)
```

# Add rectangles

Lastly, how to **add a rectangle** with the `addRectangles()` function. You have to give the **2 coordinates of 2 opposites** corners of the rectangle.

```{r}
#library
library(leaflet)
 
# Create 20 markers (Random points)
data = data.frame(
   long=sample(seq(-150,150),20),
   lat=sample(seq(-50,50),20),
   val=round(rnorm(20),2),
   name=paste("point",letters[1:20],sep="_")
) 
 
# Show a rectangle
m = leaflet() %>% addTiles() %>%  
  addProviderTiles("Stadia.StamenTonerLite") %>% 
  addRectangles(
    lng1=-118.456554, lat1=34.078039,
    lng2=-118.436383, lat2=34.062717,
    fillColor = "transparent"
  )
m

# save the widget
# library(htmlwidgets)
# saveWidget(m, file="HtmlWidget/leaflet-182-add-rectangles.html", selfcontained = F)
```

# Loading a tile

This post follows the previous posts on `leaflet`: most [basic map](https://r-graph-gallery.com/179-show-a-map-with-leaflet-r.html), and zooming on a [specific area](https://r-graph-gallery.com/179-2-select-a-zone-in-leaflet-map.html).

`Leaflet` offers several tiles to customize the background used on a map.

A tile can be loaded thanks to the `addProviderTiles()` function.

```{r}
# Load the library
library(leaflet)

# Note: if you do not already installed it, install it with:
# install.packages("leaflet")

# Background 1: NASA
m <- leaflet() %>% 
   addTiles() %>% 
   setView( lng = 2.34, lat = 48.85, zoom = 5 ) %>% 
   addProviderTiles("NASAGIBS.ViirsEarthAtNight2012")
m
 
# Background 2: World Imagery
m <- leaflet() %>% 
   addTiles() %>% 
   setView( lng = 2.34, lat = 48.85, zoom = 3 ) %>% 
   addProviderTiles("Esri.WorldImagery")
m

# costa rica
m_cr <- leaflet() %>% 
  addProviderTiles("NASAGIBS.ViirsEarthAtNight2012") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_sf10 <- leaflet() %>% 
  addProviderTiles("NASAGIBS.ViirsEarthAtNight2012") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_sf10

# save the widget in a html file if needed.
# library(htmlwidgets)
# saveWidget(m, file=paste0( getwd(), "/HtmlWidget/backgroundMapTile.html", width="1000px"))
```

# List of tiles

Many tiles are made available. A complete list can be found [here](http://leaflet-extras.github.io/leaflet-providers/preview/index.html). Most famous are probably:

- Nasa: `NASAGIBS.ViirsEarthAtNight2012`

- Google map: `Esri.WorldImagery`

- Gray: `Esri.WorldGrayCanvas`

- Terrain: `Esri.WorldTerrain`

- Topo Map: `Esri.WorldTopoMap`

```{r}
# costa rica
m_cr <- leaflet() %>% 
  addProviderTiles("NASAGIBS.ViirsEarthAtNight2012") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldGrayCanvas") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldTerrain") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldTopoMap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerBlacklite") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Stadia.StamenTonerDark") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Stadia.StamenWatercolor") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Thunderforest.OpenCycleMap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Thunderforest.SpinalMap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Thunderforest.MobileAtlas") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Thunderforest.Neighbourhood") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("CyclOSM") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr
```

more tiles

```{r}
m_cr <- leaflet() %>% 
  addProviderTiles("Jawg.Matrix") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.WorldStreetMap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("Esri.OceanBasemap") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("CartoDB.PositronNoLabels") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("CartoDB.DarkMatter") %>% 
  setView(lng = -83.7534, lat = 9.7489, zoom = 7)

m_cr
```

more

```{r}
m_cr <- leaflet() %>% 
  addProviderTiles("OpenRailwayMap") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("WaymarkedTrails.skating") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("OpenRailwayMap") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("SafeCast") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("WaymarkedTrails.cycling") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr

m_cr <- leaflet() %>% 
  addProviderTiles("WaymarkedTrails.hiking") %>% 
  setView(lng = -122.4419, lat = 37.7664, zoom = 9)

m_cr
```

# The `Maps`, `Mapdata` and `Oz` packages to get most common boundaries

The [maps](https://r-graph-gallery.com/278-the-maps-library.html#maps), [mapdata](https://r-graph-gallery.com/278-the-maps-library.html#mapdata) and [oz](https://r-graph-gallery.com/278-the-maps-library.html#oz) packages provide the boundaries of the most common world regions like the US, Europe, New Zealand and more. This post describes how to use them to avoid struggling finding input data.

# `maps` package

The [maps](https://cran.r-project.org/web/packages/maps/maps.pdf) package is the best source of geospatial data in R.

The whole list of offered data is available typing: `help(package='maps')`.

It includes:

- World: `world`, `world.cities`, `lakes`

- US: states, `county`, `state`, `usa`

- France: `france`

- Italy: `italy`

- New zealand: `nz`

The code below shows how to use one of this geospatial object.

```{r}
help(package='maps')
```

```{r}
# Load library
library(maps)

# Check all available geospatial objects:
# help(package='maps')

# Map of the world:
map('world',col="grey", fill=TRUE, bg="white", lwd=0.05, mar=rep(0,4),border=0, ylim=c(-80,80) )
```

# `mapdata` package

The [mapdata](https://cran.r-project.org/web/packages/mapdata/mapdata.pdf) package extends the `maps` package with more geospatial datasets:

- `china`

- `japan`

- Other world versions like pacific Centric (`world2Hires`)

See complete list with `help(package='mapdata')`

```{r}
help(package='mapdata')
```

```{r}
# Load library
library(mapdata)

# Check all available geospatial objects:
# help(package='mapdata')

# Map of Japan:
map('japan',col="black", lwd=1, mar=rep(0,4) )
```

```{r}
# Load library
library(mapdata)

# Check all available geospatial objects:
# help(package='mapdata')

# Map of Japan:
map('china',col="black", lwd=1, mar=rep(0,4) )
```

# `oz` package

The [oz](https://github.com/cran/oz) package provides several geospatial object concerning Australia.

Plot the general country using the `oz()` function, or states using `nsw()`, `qld()` and so on..

Type `help(package='oz')` for documentation.

```{r}
install.packages('oz')
```

```{r}
help(package='oz')
```

```{r}
# Load library
library(oz)

# Check all available geospatial objects:
# help(package='oz')

# Map of Australia
par(mar=rep(0,4))
oz( states=TRUE, col="#69b3a2")
```
