---
title: "ConnectingRoutes"
format: html
editor: visual
---

# How to draw connecting routes on map with R

This post explains how to draw connection lines between several locations on a map, using `R`. Method relies on the `gcIntermediate` function from the `geosphere` package. Instead of making straight lines, it draws the shortest routes, using [great circles](https://en.wikipedia.org/wiki/Great_circle).

# Draw an empty map

Let’s start by drawing a world map backgroud. Fortunately, the world shape is provided by the `maps` library. It can easily be plotted using the `map()` function that accepts a few arguments for customization.

[Note]{.underline}: visit the [background map](https://r-graph-gallery.com/map.html) section if you want to discover more options to build map in `R`.

```{r}
# World map is available in the maps package
library(maps)

# No margin
par(mar=c(0,0,0,0))

# World map
map('world',
    col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05,
    mar=rep(0,4),border=0, ylim=c(-80,80) 
)
```

# Add 3 cities

It’s relatively straightforward to add circles at specific positions using the `points()` function. The hardest part is probably to find out the longitute and latitude of the cities you are interested in.

```{r}
# Dplyr for data wrangling and pipe function
library(dplyr)

# Cities
Buenos_aires <- c(-58,-34)
Paris <- c(2,49)
Melbourne <- c(145,-38)

# Data frame
data <- rbind(Buenos_aires, Paris, Melbourne) %>% 
  as.data.frame()
colnames(data) <- c("long","lat")

# Show the cities on the map
map('world',
    col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05,
    mar=rep(0,4),border=0, ylim=c(-80,80) 
)
points(x=data$long, y=data$lat, col="slateblue", cex=3, pch=20)
```

# Show connection between cities

Now we can connect cities drawing the shortest route between them. This is done using great circles, what gives a better visualization than using straight lines.

The `geosphere` package provides the `gcIntermediate()` function that computes the path. This path can then be plotted using the `lines()` function.

```{r}
# Load geosphere
library(geosphere)

# Background map
map('world',
    col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05,
    mar=rep(0,4),border=0, ylim=c(-80,80) 
)

# Dot for cities
points(x=data$long, y=data$lat, col="slateblue", cex=3, pch=20)

# Compute the connection between Buenos Aires and Paris
inter <- gcIntermediate(Paris,  Buenos_aires, n=50, addStartEnd=TRUE, breakAtDateLine=F)

# Show this connection
lines(inter, col="slateblue", lwd=2)
 
# Between Paris and Melbourne
inter <- gcIntermediate(Melbourne,  Paris, n=50, addStartEnd=TRUE, breakAtDateLine=F)             
lines(inter, col="slateblue", lwd=2)
```

Using the exact same method between Melbourne and Buenos Aires gives a pretty disapointing result as shown here.

Indeed, there is a long horizontal line that appears, when we expect the connection to be done on the other side of the globe (and thus invisible)

What happens is that `gcintermediate` follows the shortest path, which means it will go east from Australia until the date line, break the line and come back heading East from the pacific to South America.

Because we do not want to see the horizontal line, we need to plot this connection in 2 steps. To do so we can use the following function, which breaks the line in 2 sections when the distance between 2 points is longer than 180 degrees:

```{r}
library(maps)
library(geosphere)

# 1. Define city coordinates (Longitude, Latitude)
Paris        <- c(2.35, 48.85)
Buenos_aires <- c(-58.38, -34.60)
Melbourne    <- c(144.96, -37.81)

# Create a data frame for the city dots
data <- data.frame(
  long = c(Paris[1], Buenos_aires[1], Melbourne[1]),
  lat  = c(Paris[2], Buenos_aires[2], Melbourne[2])
)

# 2. Draw the Background map
# We use par(mar...) to handle margins instead of putting 'mar' inside map()
par(mar=c(0,0,0,0))
map('world',
    col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05,
    border=NA, ylim=c(-80,80) 
)

# 3. A robust function to plot connections
# This handles the "wrap around" for the dateline automatically
plot_my_connection <- function(dep_lon, dep_lat, arr_lon, arr_lat, ...){
    inter <- gcIntermediate(c(dep_lon, dep_lat), c(arr_lon, arr_lat), 
                            n=50, addStartEnd=TRUE, breakAtDateLine=TRUE)             
    
    # If it's a list (happens when breakAtDateLine=TRUE), plot each segment
    if(is.list(inter)){
      lines(inter[[1]], ...)
      lines(inter[[2]], ...)
    } else {
      lines(inter, ...)
    }
}

# 4. Draw the connections using the function
plot_my_connection(Paris[1], Paris[2], Buenos_aires[1], Buenos_aires[2], col="slateblue", lwd=2)
plot_my_connection(Melbourne[1], Melbourne[2], Paris[1], Paris[2], col="slateblue", lwd=2)
plot_my_connection(Buenos_aires[1], Buenos_aires[2], Melbourne[1], Melbourne[2], col="slateblue", lwd=2)

# 5. Add the dots for cities on top
points(x=data$long, y=data$lat, col="slateblue", cex=2, pch=20)
```

# Apply it to several pairs of cities

If we want to plot many connections, it is totally doable to build a `for` loop that draws them one by one.

Moreover, note the use of the `text` function to add city names.

```{r}

# Load libraries
library(dplyr)
library(maps)
library(geosphere)

# 1. Create dataset (Corrected to: Longitude, Latitude)
data <- data.frame(
    long = c(18.4, 2.35, 28.0, 103.8, 115.2, 139.6, -99.1, -61.4, -157.8, -122.4, -74.0, 12.3),
    lat = c(-33.9, 48.8, -26.2, 1.3, -8.4, 35.6, 19.4, 12.5, 21.3, 37.8, 40.7, 45.4),
    row.names = c("Cape_town", "Paris", "Johannesburg", "Singapore", "Bali", "Tokyo", 
                  "Mexico_city", "Carriacou", "Honolulu", "San_Francisco", "New_York_City", "Venice")
)

# 2. Re-define the robust connection function
plot_my_connection <- function(dep_lon, dep_lat, arr_lon, arr_lat, ...){
    inter <- gcIntermediate(c(dep_lon, dep_lat), c(arr_lon, arr_lat), 
                            n=50, addStartEnd=TRUE, breakAtDateLine=TRUE)             
    if(is.list(inter)){
      lines(inter[[1]], ...)
      lines(inter[[2]], ...)
    } else {
      lines(inter, ...)
    }
}

# 3. Background map
par(mar=c(0,0,0,0))
map('world', col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05, border=NA, ylim=c(-80,80))

# 4. Generate pairs by INDEX to keep Lat/Long together
pairs <- combn(1:nrow(data), 2)

# 5. Loop through pairs
for(i in 1:ncol(pairs)){
    p1 <- data[pairs[1, i], ]
    p2 <- data[pairs[2, i], ]
    plot_my_connection(p1$long, p1$lat, p2$long, p2$lat, col="skyblue", lwd=0.5)
}

# 6. Add points and names
points(x=data$long, y=data$lat, col="slateblue", cex=1, pch=20)
text(rownames(data), x=data$long, y=data$lat,  col="slateblue", cex=0.7, pos=4)
```

# An alternative using the greatCircle function

This is the method suggested by the [Simply Statistics](http://simplystatistics.tumblr.com/post/14568185310/an-r-function-to-map-your-twitter-followers) Blog to draw a twitter connection map.

The idea is to calculate the whole great circle, and keep only the part that stays in front of the map, never going behind it.

```{r}
library(maps)
library(geosphere)

# 1. Define your cities (Longitude, Latitude)
# Make sure these are defined before you call them!
Paris     <- c(2.35, 48.85)
Honolulu  <- c(-157.85, 21.3)
Tokyo     <- c(139.6, 35.6)
Cape_town <- c(18.4, -33.9)

# Put them in a data frame for the dots/labels
data <- data.frame(
  long = c(Paris[1], Honolulu[1], Tokyo[1], Cape_town[1]),
  lat  = c(Paris[2], Honolulu[2], Tokyo[2], Cape_town[2]),
  row.names = c("Paris", "Honolulu", "Tokyo", "Cape_town")
)

# 2. Draw the background map
par(mar=c(0,0,0,0))
map('world', col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05, border=NA, ylim=c(-80,80))

# 3. Use gcIntermediate for reliable lines
# We wrap it in a helper to handle the Date Line split automatically
draw_line <- function(start, end) {
  inter <- gcIntermediate(start, end, n=50, addStartEnd=TRUE, breakAtDateLine=TRUE)
  if (is.list(inter)) {
    lines(inter[[1]], col="skyblue", lwd=2)
    lines(inter[[2]], col="skyblue", lwd=2)
  } else {
    lines(inter, col="skyblue", lwd=2)
  }
}

draw_line(Paris, Honolulu)
draw_line(Honolulu, Tokyo)
draw_line(Paris, Cape_town)

# 4. Add dots and text
points(x=data$long, y=data$lat, col="slateblue", cex=2, pch=20)
text(x=data$long, y=data$lat, labels=rownames(data), col="slateblue", cex=0.8, pos=4)
```
