---
title: "Hierarchical cluster dendrogram with hclust"
format: html
editor: visual
---

# **Hierarchical cluster dendrogram with hclust**

## **Sample data set**

The following data set contains statistics about arrests in the USA. For illustration purpose we are going to use the first twenty rows of the data set in the following examples.

```{r}
df <- USArrests[1:20, ]
df
```

## **Basic dendrogram**

In order to create a dendrogram in R first you will need to calculate the distance matrix of your data with `dist`, then compute the hierarchical clustering of the distance matrix with `hclust` and plot the dendrogram.

**Option 1**

Plot the hierarchical clustering object with the `plot` function.

```{r}
# Distance matrix
d <- dist(df)

# Hierarchical clustering
hc <- hclust(d)

# Dendrogram
plot(hc)
```

**Option 2**

Transform the hierarchical clustering output to dendrogram class with `as.dendrogram`. This will create a nicer visualization.

```{r}
# Distance matrix
d <- dist(df)

# Hierarchical clustering
hc <- hclust(d)

# Dendrogram
plot(as.dendrogram(hc))

# The previous line is similar to:
# plot(hc, hang = -1)
```

The `dist` function computes the distance matrix based on the euclidean distance, but there are other distance measures available such as `"maximum"`, `"manhattan"`, `"canberra"`, `"binary"` or `"minkowski"`. Choose the one that best fits your data.

## **Clustering methods**

In addition to several distance measures there are several hierarchical clustering methods you can choose. The default method is `"complete"`, but you can choose between `"ward.D"`, `"ward.D2"`, `"single"`, `"average"`, `"mcquitty"`, `"median"` or `"centroid"`. Type `?hclust` for additional details about each method and other arguments.

In the examples below you can see the difference using each method.

```{r}
# Distance matrix
d <- dist(df)

# Hierarchical clustering
hc <- hclust(d, method = "ward.D")

# Dendrogram
plot(as.dendrogram(hc), main = "ward.D")
```

```{r}
# Distance matrix
d <- dist(df)

# Hierarchical clustering
hc <- hclust(d, method = "ward.D2")

# Dendrogram
plot(as.dendrogram(hc), main = "ward.D2")
```

```{r}
# Distance matrix
d <- dist(df)

# Hierarchical clustering
hc <- hclust(d, method = "single")

# Dendrogram
plot(as.dendrogram(hc), main = "single")
```
