---
title: "Area chart in ggplot2 with geom_area"
format: html
editor: visual
---

# **Area chart in ggplot2 with geom_area**

## **Sample data set**

Consider the following data frame that contains the price of three european stock indices.

```{r}
df <- as.data.frame(EuStockMarkets[, 1:3])
df
```

## **Area chart of a single line**

In order to create an area chart with a column of the input data frame you can pass the dates (if available) or an index to `x` and the variable of interest to `y` and then use `geom_area`.

```{r}
# install.packages("ggplot2")
library(ggplot2)

# Area chart
ggplot(df, aes(x = seq_along(SMI), y = SMI)) + 
  geom_area() +
  custom_theme_white_nolegend()
```

**Fill the area**

By default the area will be filled with a very dark gray color. However, you can change it passing a color to the `fill` argument.

```{r}
# install.packages("ggplot2")
library(ggplot2)

# Area chart
ggplot(df, aes(x = seq_along(SMI), y = SMI)) + 
  geom_area(fill = 4)  +
  custom_theme_black_nolegend()

```

**Transparency of the area**

You can also change the transparency of the area with the `alpha` argument of the function.

```{r}
# install.packages("ggplot2")
library(ggplot2)

# Area chart
ggplot(df, aes(x = seq_along(SMI), y = SMI)) + 
  geom_area(fill = 4,
            alpha = 0.5)  +
  custom_theme_black_nolegend()

```

**Line customization**

The upper line of the area can be customized with several arguments, such as `color`, `lwd` or `linetype`.

```{r}
# install.packages("ggplot2")
library(ggplot2)

# Area chart
ggplot(df, aes(x = seq_along(SMI), y = SMI)) + 
  geom_area(fill = 4,
            alpha = 0.5,
            color = 1,    # Line color
            lwd = 0.5,    # Line width
            linetype = 1) + # Line type 
  custom_theme_white_nolegend()
 
 

```

## **Area chart of several lines**

In case you want to add more variables you will need to specify the aesthetics inside each `geom_area` for each variable. Note that if the number of areas to add is more than two or three you should consider creating a stacked area chart.

```{r}
# install.packages("ggplot2")
library(ggplot2)

# Area chart
ggplot(df) + 
  geom_area(aes(x = seq_along(SMI), y = SMI),
            fill = 4, alpha = 0.85) +
  geom_area(aes(x = seq_along(DAX), y = DAX),
            fill = 3, alpha = 0.85) +
  custom_theme_black_nolegend()
```
