---
title: "Circular dendrogram in R with circlize and dendexten"
format: html
editor: visual
---

# **Circular dendrogram in R with circlize and dendexten**

## **Circular fan (polar) dendrogram plot**

The `circlize_dendrogram` function can be used to create a circular [dendrogram](https://r-charts.com/part-whole/hclust/), also known as circular, radial or fan tree plot. You need to pass an object of class dendrogram. Note that `labels_track_height` and `dend_track_height` arguments are used to fit the labels properly and that both `circlize` and `dendextend` need to be installed to run the function.

```{r}
install.packages("dendextend")
install.packages("circlize")
```

```{r}
# install.packages("dendextend")
# install.packages("circlize")
library(dendextend)
library(circlize)

# Distance matrix
d <- dist(USArrests)

# Hierarchical clustering dendrogram
hc <- as.dendrogram(hclust(d))

# Circular dendrogram
circlize_dendrogram(hc,
                    labels_track_height = NA,
                    dend_track_height = 0.5) 
 
 

```

The default dendrogram faces outside, but you can change this setting `facing = "inside"`.

```{r}
# install.packages("dendextend")  
# install.packages("circlize")
library(dendextend)
library(circlize)

# Distance matrix
d <- dist(USArrests)

# Hierarchical clustering dendrogram
hc <- as.dendrogram(hclust(d))

# Circular dendrogram facing inside
circlize_dendrogram(hc,
                    labels_track_height = NA,
                    dend_track_height = 0.3,
                    facing = "inside",
                    width = 1200,  # Adjust width
                    height = 1200)  # Adjust height

```

The labels of the dendrogram can be removed if you only want to display the branches setting `labels = FALSE`. In this scenario is recommended to increase the `dend_track_height` argument value.

```{r}
# install.packages("dendextend")
# install.packages("circlize")
library(dendextend)
library(circlize)

# Distance matrix
d <- dist(USArrests)

# Hierarchical clustering dendrogram
hc <- as.dendrogram(hclust(d))

# Circular dendrogram without labels
circlize_dendrogram(hc,
                    dend_track_height = 0.8,
                    labels = FALSE) 

```

## **Further customization**

The styling of the polar tree plot can be customized transforming the dendrogram object. If you want to **colorize the branches** you can use the `color_branches` function, based on a number of clusters (`k`) or a height (`h`). Note that the function contains an argument named `col` that allows customizing the colors.

```{r}
# install.packages("dendextend")
# install.packages("circlize")
library(dendextend)
library(circlize)

# Distance matrix
d <- dist(USArrests)

# Hierarchical clustering dendrogram
hc <- as.dendrogram(hclust(d))

# Colors
hc <- hc %>%
  color_branches(k = 3)
# hc <- color_branches(hc, k = 4)

# Radial plot
circlize_dendrogram(hc,
                    labels_track_height = NA,
                    dend_track_height = 0.5,
                    width = 1200,  # Increase width
                    height = 1200,  # Adjust height
                    circular = TRUE,  # Ensure circular layout
                    rotation = 90)  # Rotate dendrogram for better fit

```

The **labels can also be customized** the same way as the branches. For that purpose use the `color_labels` function and specify the number of clusters or the height.

```{r}
# install.packages("dendextend")
# install.packages("circlize")
library(dendextend)
library(circlize)

# Distance matrix
d <- dist(USArrests)

# Hierarchical clustering dendrogram
hc <- as.dendrogram(hclust(d))

# Colors
hc <- hc %>%
  color_branches(k = 3) %>%
  color_labels(k = 3)

# Fan tree plot with colored labels
circlize_dendrogram(hc,
                    labels_track_height = NA,
                    dend_track_height = 0.5)   

```

Finally, it is worth to mention that the **style and width of the lines** can be changed as in the example below.

```{r}
# install.packages("dendextend")
# install.packages("circlize")
library(dendextend)
library(circlize)

# Distance matrix
d <- dist(USArrests)

# Hierarchical clustering dendrogram
hc <- as.dendrogram(hclust(d))

# Colors and line customization
hc <- hc %>%
  color_branches(k = 3) %>%
  set("branches_lwd", 2) %>%  # Line width
  set("branches_lty", 2) %>%  # Line type
  color_labels(k = 3)

# Line styling of the dendrogram
circlize_dendrogram(hc,
                    labels_track_height = NA,
                    dend_track_height = 0.5)   
 

```

Example 20 rows

```{r}
# install.packages("dendextend")
# install.packages("circlize")
library(dendextend)
library(circlize)

# Distance matrix
# Subset the first 20 rows from the USArrests dataset
subset_data <- USArrests[1:10, ]

# Calculate the distance matrix
d2 <- dist(subset_data)


# Hierarchical clustering dendrogram
hc <- as.dendrogram(hclust(d2))

# Circular dendrogram
circlize_dendrogram(hc,
                    labels_track_height = NA,
                    dend_track_height = 0.5) 
```
