---
title: "Simple Dashboard in R"
format: html
editor: visual
---
### Not working yet

## Creating a simple dashboard in R with shiny

Yes, it is possible to create a dashboard in R using various packages, and one popular choice is the **`shiny`** package. Shiny is an R package that makes it easy to build interactive web applications directly from R.

Here's a simple example of creating a basic dashboard with **`shiny`**:

1.  First, make sure you have the **`shiny`** package installed. If you don't have it installed, you can install it by running:

```{r}
install.packages("shiny")

```

Create a new R script or RMarkdown document, and let's build a simple Shiny dashboard. Here's an example:

```{r}
# Install and load the required libraries
if (!require(shiny)) {
  install.packages("shiny")
}
library(shiny)

# Define the UI
ui <- fluidPage(
  titlePanel("Simple Shiny Dashboard"),
  sidebarLayout(
    sidebarPanel(
      textInput("name", "Enter your name:", ""),
      sliderInput("age", "Select your age:", 1, 100, 25),
      actionButton("submit", "Submit")
    ),
    mainPanel(
      textOutput("output")
    )
  )
)

# Define the server logic
server <- function(input, output) {
  observeEvent(input$submit, {
    output$output <- renderText({
      paste("Hello, ", input$name, "! You are ", input$age, " years old.")
    })
  })
}

# Run the application
shinyApp(ui, server)

```

1.  In this example:

    -   The UI is defined using the **`fluidPage`** function, containing a title panel, a sidebar layout with a sidebar panel, and a main panel.

    -   The server logic is defined using the **`server`** function, which reacts to the submit button click event and updates the output accordingly.

    -   The **`shinyApp`** function is used to launch the Shiny application.

2.  Save the script and run it in your R environment.

3.  The Shiny application will create a simple dashboard where users can enter their name, age, and click the submit button to see a personalized message.

This is a basic example, and Shiny allows you to create much more complex and sophisticated dashboards with interactive plots, data tables, and more. You can customize the appearance and functionality based on your needs.
