---
title: "Spineplot in R"
format: html
editor: visual
---

# **Spineplot in R**

## **Data**

The following table contains the count the number of Titanic survivals based of their economic status (class). This data will be used in the examples below.

```{r}
# Data 
dat <- apply(Titanic, c(1, 4), sum)

dat
```

## **Spineplot**

Spineplots are a generalization of [stacked bar charts](https://r-charts.com/part-whole/stacked-bar-graph/). The `spineplot` function allows creating this type of graph in base R (actually a mosaic plot, which is a spineplot for two or more variables). See `?spineplot` to know more about the possible input options.

```{r}
spineplot(dat)
```

**Transpose the variables**

The spineplot can be flipped transposing the data, as in the example below.

```{r}
spineplot(t(dat))
```

### **Color customization**

**Fill color**

You can pass a vector of colors to `col` to customize the default gray scale colors. Pass as many colors as classes on the Y-axis.

```{r}
spineplot(dat, col = c("#07798D", "#7BCB9F"))
```

**Border color**

You can also customize the border color with `border`. In this example we are setting it to the fill color.

```{r}
spineplot(dat, col = c("#07798D", "#7BCB9F"),
          border = c("#07798D", "#7BCB9F"))
```

**Spinogram**

Spinograms are a generalization of histograms. The default `spineplot` function creates a spineplot, but setting `off = 0` a spinogram will be created, removing the offset between the bars.

```{r}
spineplot(dat, off = 0,
          col = c("lightgreen", "red"),
          border = c("white", "white"))
```
